Oriented Gradient Histogram applied to P300 Detection

### 2017-11-27 response.letter.tex

\documentclass[journal,onecolumn,12pt]{IEEEtran}

\usepackage{amsmath,amssymb,bm}
\usepackage{amsthm, amsfonts}	
\usepackage{bm,bbm}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{fancybox}
\usepackage{url,booktabs}

\usepackage{xr}

\title{Reply to Reviewer's Comments on\\
``EEG characterization and classification based on image gradient histograms''\\
(now ``Oriented Gradient Histogram applied to P300 Detection'')}
\author{}

\begin{document}

\maketitle
\pagenumbering{roman}
\setcounter{page}{1}

We are grateful to the reviewers for pointing at relevant issues in our manuscript.

In the following, we discuss how we dealt with each raised issue.

\vskip+1ex
\noindent \dotfill

\section*{\fbox{Reviewer \#1:}}

I have read the manuscript "EEG characterization and classification based on Histogram of Gradients of Signal Plots" with great interest since its title promised a solution to an important problem. In spite of presenting an interesting methodology and some interesting issues related to EEG characterization, I consider the manuscript is not ready for publication.

First of all, the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few). Clearly, the title does not match the content since the authors focus on the P300; of course, there is evidence of the benefits of working with these signals, but, is the proposed method extensible to other signals?, this is not clear from the manuscript.

Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors. Moreover, there is something wrong in equation (4) and something inconsistent between equations (7) and (8). Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.

Another issue that I find worrisome is the design of experiments. First of all, it seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset. Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.

A couple of finale observations. First, it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference. Another issue has to do with the way the manuscript is written, sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5), the manuscript surely benefit from a careful review before submitting.

\subsection*{\ovalbox{General Comments}}

2.1) the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few). Clearly, the title does not match the content since the authors focus on the P300

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. You're very clever.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2.2) these signals, but, is the proposed method extensible to other signals?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
This comment is very important, thank you very much. 
An explanation  of which is a satisfactory result for Algorithm~\ref{Algoritmo} was added in sections~\ref{errorevaluation} and~\ref{sec:EdgeSimulatedData}.  
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
\subsection*{\ovalbox{Third Paragraph}}

3.1) Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  You are so clever that it is very likely that you are God.}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.2) Moreover, there is something wrong in equation (4)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Oh my God, I can't believe how smart you are !!! You probably are rich by now.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.3) and something inconsistent between equations (7) and (8)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have revised the manuscript and made a clear distinction between ``methods'' and ``features''.
We thank you for noticing this, that led to a less ambiguous text.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.4) Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  Smart people like you should be president.
Following your suggestion, we added graphical evidence of the difficulty of separating targets characterized by $\mathcal{G}_I^0(-2,1,1)$, $\mathcal{G}_I^0(-3,1,1)$, $\mathcal{G}_I^0(-4,1,1)$, and $\mathcal{G}_I^0(-5,1,1)$ data; cf.\ Fig.~\ref{fig:DifficultProblem}.
With this, and the corresponding discussion, we believe we have addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

\subsection*{\ovalbox{Fourth Paragraph}}

4.1) Another issue that I find worrisome is the design of experiments.\\

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. 
We now specify details of the E-SAR image used in the applications to true data; cf. Sec.~\ref{sec:ActualData}.
Following your suggestion, we added  a comparison with the Triangular distance in actual data. 
We applied the method in edge detection with both distances and we discovered that both of them resulted in exactly the same edge point. 
Then, we concluded that these distances are very similar in terms of accuracy in finding the transition point, but the geodesic distance is better in terms of  computation time. 
We fully addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

4.2) it seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your suggestion. We increased the discussion and conclusions, and enhanced lines for future research.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

4.3) Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Oh God. You are clearly an oracle. 
All these points were modified according to the suggestions, but instead of doing what you asked for we do what we want instead.
We hope this is acceptable.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

\subsection*{\ovalbox{Fifth Paragraph}}

5.1) it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Oh God. You are clearly an oracle. 
All these points were modified according to the suggestions, but instead of doing what you asked for we do what we want instead.
We hope this is acceptable.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

5.2)  issue has to do with the way the manuscript is written

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Oh God. You are clearly an oracle. 
All these points were modified according to the suggestions, but instead of doing what you asked for we do what we want instead.
We hope this is acceptable.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

5.3) sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Oh God. You are clearly an oracle.

We hope this is acceptable.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

\end{document}

### 2017-11-27 article.tex
First set of changes for the Resubmission

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG characterization and classification based on Histogram of Gradients of Signal Plots}

@@

Moreover, EEG markers \citep{Clerc} that can be used to volitionally transmit information are limited, and each one of them has a particular combination of appropriate methods to decode them. Inevitably, it is necessary to implement many distinct and specialized algorithmic methods, to filter the signal, enhance its SNR, and try to determine some meaning out of it.

@@

The P300 \citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs 300 ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm \cite{WolpawJonathanR2012} and though it is quite consistent across different subjects, it has a lower amplitude ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around -15 db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity \citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix \citep{Farwell1988} (Fig. \ref{fig:p300matrix}) where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

@@

The first step consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. 
The processing pipeline starts by applying a notch filter to the raw signal, a 
4th degree 10 Hz lowpass Butterworth filter and finally a decimation with an FIR filter of order 30 from the original 256 Hz down to 16 Hz \citep{Krusienski2006}.

@@

\begin{itemize}
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the 6 rows plus the 6 columns, a basic artefact elimination procedure is implemented by removing the entire segment when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the 12 stimuli,  a window of 1 second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments are labeled as \textit{hit}, whereas the remaining 10 are labeled as \textit{no hit}.  A hit represents that the EEG segment should contain the ERP signature time-locked to the flashing stimulus. 
\item \textbf{Signal Averaging}: The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

The underlying idea of this method is to generate a template for the signal,  based on the image plot.  Hence, the first step is its transformation into a temporary binary image.

@@

In order to complete the plot from the pixels, the Bresenham \citep{Ramele2016} algorithm is used to interpolate straight lines between each pair of these consecutive pixels.

@@

On the generated image, a keypoint is placed on a precise location along the signal trace over the image plot (Fig. \ref{fig:sampledescriptor}).  Around that keypoint, a local image patch is constructed as follows:  it is first divided in 16 blocks of size $ 3 \cdot s $ each, arranged in a 4x4 grid and centered on $T = (x,y)$.

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the 16 blocks and creating a histogram of the such gradients.  This technique is based on Lowe's SIFT \citep{Lowe2004} descriptor, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate this histogram, the interval $ [0-360]$ of possible angles is divided in 8 bins, each one at 45 degrees.

@@

\begin{equation}
 h(p,i,j) = 3 \cdot s \int w_\mathrm{ang}(\angle J(\mathbf{x}) - \theta_p)\, w_{ij}\left(\frac{\mathbf{x} - T}{3 s}\right)\, |J(\mathbf{x})|\, d\mathbf{x}\, d\theta_p
\label{eq:histogram}
\end{equation}

\noindent where $s$ is the size of the local patch, which can be converted into pixels by doing $ Px = 4 \cdot 3 \cdot s $ ,  $ |J(\mathbf{x})| $ is the norm of the gradient vector found at each one of the 16 blocks of the patch, whereas $\angle J(\mathbf{x}) $ is the angle of the gradient vector and  $\theta_p$ is the angle bin.  On the other hand $ \textbf{x} = (x_i, x_j) $ corresponds to each one of the pixels of the local patch, and $ w_\mathrm{ang}  $  and $ w_{ij} $ are linear interpolation functions \citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each grid block when the size of the patch is the unity.  As the patch is 4x4, it gives a descriptor of 128 dimension as shown on Figure \ref{fig:sampledescriptor}.

\begin{figure}[H]
\centering
\includegraphics[width=6cm]{sampledescriptor.png}
\includegraphics[width=9cm]{histogramsbines.png}
\caption{(Left) The local patch is located on a precise location within the generated plot. The 4x4 block grid can be seen on the patch. (Right) On each grid block, eight bins are used to describe the eight different orientations, and they are later arranged in a 128-dimension vector (4x4x8).}
\label{fig:sampledescriptor}
\end{figure}

@@

The experimental protocol used to generate this dataset is explained in \citep{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The P300 detection task designed for this experiment consisted of spelling 7 words (runs) of 5 letter each, using the traditional P300 Speller Matrix \citep{Farwell1988} where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.  The first 3 runs were used for training and the remaining 4 for testing with visual feedback.  A trial, as defined by the BCI2000 platform \citep{Schalk2004}, was every attempt to select a letter from the speller, and it was composed of signal segments corresponding to 10 repetitions of flashes of 6 rows and 6 columns of the traditional 6x6 P300 matrix , yielding 120 repetitions.  Flashing of row/columns was performed for 0.125 s, following by a resting period (i.e. Inter-stimulus interval) of the same length.  After 120 repetitions an inter-trial pause was included before resuming with the following letter.

@@

\begin{equation}
s = \frac{\Delta \mu V}{4 \cdot 3} \cdot \gamma 
\label{eq:mapping1}
\end{equation}

@@

%\begin{figure}[H]
%\centering
%\includegraphics[width=16cm]{localpatch.png}
%\caption{The local patch is mapped to the original image by their correspondence .}
%\label{fig:sampledescriptor}
%\end{figure}

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Oriented Gradient Histogram applied to P300 Detection}

@@

Moreover, EEG markers \citep{Clerc} that can be used to volitionally transmit information are limited, and each one of them has a particular combination of appropriate methods to decode them. Inevitably, it is necessary to implement  distinct and specialized algorithmic methods, to filter the signal, enhance its Signal to Noise Ratio (SNR), and try to determine some meaning out of it.

@@

The P300 \citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs around 300 ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm \cite{WolpawJonathanR2012} and though it is quite consistent across different subjects, it has a lower amplitude  ($\pm 5 \mu V $) compared to basal EEG activity, reaching a SNR of around -15 db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity \citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix \citep{Farwell1988} as shown in Fig. \ref{fig:p300matrix}, where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

@@

The first step consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. 
The processing pipeline starts by applying a notch filter to the raw signal, a 
4th degree 10 Hz lowpass Butterworth filter and finally a decimation with an Finite Impulse Response (FIR) filter of order 30 from the original sampling frequency down to 16 Hz \citep{Krusienski2006}.

@@

\begin{itemize}
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the 6 rows plus the 6 columns, a basic artefact elimination procedure is implemented by removing the entire segment when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the 12 stimuli,  a window of 1 second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments are labeled as \textit{hit}, whereas the remaining 10 are labeled as \textit{no hit}.  A hit represents that the EEG segment should contain the P300 ERP signature time-locked to the flashing stimulus. 
\item \textbf{Signal Averaging}:  The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

@@

The underlying idea of this method is to generate a template for the signal \citep{Ramele2016},  based on the image plot.  Hence, the first step is its transformation into a temporary binary image.

@@

In order to complete the plot from the pixels, the Bresenham \citep{Bresenham1965,Ramele2016} algorithm is used to interpolate straight lines between each pair of these consecutive pixels.

@@

On the generated image, a keypoint $\mathbf{kp}$ is placed on a pixel $(x_kp, y_kp)$ over the image plot. In Figure  \ref{fig:sampledescriptor}.  Around the keypoint, a local image patch is constructed by dividing it in 16 blocks of size $ 3 \cdot s $ where $ s is the size of the local patch $, arranged in a 4x4 grid and centered on $ \mathbf{kp} = (x_kp,y_kp)$, the patch center.

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the 16 blocks and creating a histogram of such gradients.  This technique is based on Lowe's SIFT \citep{Lowe2004} descriptor, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate this histogram, the interval $ [0-360]$ of possible angles is divided in 8 bins, each one at 45 degrees.

@@

\begin{equation}
 h(p,i,j) = 3 \cdot s \sum w_\mathrm{ang}(\angle J(\mathbf{x}) - \theta_p)\, w_{ij}\left(\frac{\mathbf{x} - \mathbf{kp_c}}{3 s}\right)\, |J(\mathbf{x})|
\label{eq:histogram}
\end{equation}

\noindent where $s$ is the size of the local patch, which can be converted into pixels by doing $ Px = 4 \cdot 3 \cdot s $ ,  $ |J(\mathbf{x})| $ is the norm of the gradient vector found at each one of the 16 blocks of the patch, whereas $\angle J(\mathbf{x}) $ is the angle of the gradient vector and  $\theta_p$ is the angle bin.  On the other hand $ \textbf{x} = (x_i, x_j) $ corresponds to each one of the pixels of the local patch, and $ w_\mathrm{ang}  $  and $ w_{ij} $ are linear interpolation functions \citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each grid block when the size of the patch is one (i.e. $s = 1$).  As the patch is 4x4, it gives a descriptor of 128 dimension as shown on Figure \ref{fig:sampledescriptor}.

\begin{figure}[H]
\centering
\includegraphics[width=5cm]{descriptorpatchsample.png}
\includegraphics[width=6cm]{histogramsbines.png}
%\includegraphics[width=9cm]{gradientszoom.png}
\caption{(Left) AGREGAR UN GRAFICO QUE MUESTRE MEJOR EL HISTOGRAMA Y PONER UNO DONDE EL P300 CON SU DESCRIPTOR SE VEAN PERFECTO.The local patch is located on a precise location within the generated plot. The 4x4 block grid can be seen on the patch. (Right) On each grid block, eight bins are used to describe the eight different orientations, and they are later arranged in a 128-dimension vector (4x4x8).}
\label{fig:sampledescriptor}
\end{figure}

@@

The experimental protocol used to generate this dataset is explained in \citep{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The P300 detection task designed for this experiment consisted of spelling 7 words (runs) of 5 letters each, using the traditional P300 Speller Matrix \citep{Farwell1988} where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.  The first 3 runs were used for training and the remaining 4 for testing with visual feedback.  A trial, as defined by the BCI2000 platform \citep{Schalk2004}, was every attempt to select a letter from the speller, and it was composed of signal segments corresponding to 10 repetitions of flashes of 6 rows and 6 columns of the traditional 6x6 P300 matrix , yielding 120 repetitions.  Flashing of row/columns was performed for 0.125 s, following by a resting period (i.e. Inter-stimulus interval) of the same length.  After 120 repetitions an inter-trial pause was included before resuming with the following letter.

@@

\begin{equation}
s= \frac{\Delta \mu V}{4 \cdot 3} \cdot \gamma 
\label{eq:mapping1}
\end{equation}

@@

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{localpatch.png}
\caption{The local patch captures the signal information based on both the vertical and horizontal scale.}
\label{fig:sampledescriptor2}
\end{figure}
